Model-free and Bayesian Ensembling Model-based Deep Reinforcement Learning for Particle Accelerator Control Demonstrated on the FERMI FEL

### 2025-12-29 main.tex
docs: Add `requirements.txt` and enhance `README.md` with installation instructions and improved code descriptions.

	\begin{abstract}
		Reinforcement learning holds tremendous promise in accelerator controls. The primary goal of this paper is to show how this approach can be utilised on an operational level on accelerator physics problems. Despite the success of model-free reinforcement learning in several domains, sample-efficiency still is a bottle-neck, which might be encompassed by model-based methods.
		We compare well-suited purely model-based to model-free reinforcement learning applied to the intensity optimisation on the FERMI FEL system. We find that the model-based approach demonstrates higher representational power and sample-efficiency, while the asymptotic performance of the model-free method is slightly superior. The model-based algorithm is implemented in a DYNA-style using an uncertainty aware model, and the model-free algorithm is based on tailored deep Q-learning. In both cases, the algorithms were implemented in a way, which presents increased noise robustness as omnipresent in accelerator control problems. Code is released in \url{https://github.com/MathPhysSim/FERMI_RL_Paper}.

@@

	In particle accelerator operation, one main goal is to provide stable and reproducible performance. Achieving this demands the consideration of several control problems simultaneously \cite{Kain2020,Scheinker2018}. Especially if there is no way to model the physics a priori, one might use optimisation techniques as, e.g. derivative-free optimisers (DFOs) \cite{Huang2013,Bruchon2017,Scheinker2020,Hirlaender2019,Welsch2015,Albright2019} or model-based optimisations as Gaussian processes \cite{Hanuka2020,Roussel2020} to restore or maintain the performance.\\

@@

		\item It memorises the problem and does not always begin from zero as a DFO.

@@

	One demanding aspect using RL in real-world applications is the number of iterations needed to train a controller - the sample-efficiency, since RL methods are known to be data-hungry \cite{Sutton2018,DulacArnold2019}. A second critical aspect to be considered is the robustness of the training in a noisy data-limited environment.\\

@@

	We successfully applied two different algorithm classes, a highly sample-efficient variant of continuous Q-learning (\cref{ss:Normalized advantage function}) and a newly proposed model-based algorithm, which we call \emph{AE-DYNA} (\cref{ss:Uncertainty aware DYNA-style reinforcement learning}), to the FERMI FEL problem.\\ The methods reveal different characteristics addressing the stated critical aspects. They are generally applicable to problems showing a sample-complexity as in common accelerator control problems.

@@

	 Thus, we show the operational applicability of purely model-based reinforcement learning in an accelerator control problem.\\

@@

		\item Description of the problem set-up at FERMI

@@

	\section{The set-up of the studied problem}
	\subsection{The Physical Set-up}

@@

		\caption{A schematic view on the set-up of the FERMI FEL.}

@@

	A schematic overview of the set-up is provided in \cref{fig:schematic_FEL}.

@@

	The state $\bs$ is a four dimensional vector holding the current voltage values applied to the piezo motors, where each component lies within the normalized interval [0, 1]. Two values correspond to the two DOF of each mirror. The action $\ba$ is four dimensional, namely a delta step on the voltages at time-step $t$ $\bs_t$:

@@

	\section{Deep Reinforcement learning}
	Most commonly, two main categories of deep RL algorithms are distinguished: model-free RL (MFRL) and model-based RL (MBRL). MFRL algorithms learn directly from the interaction with a system, while MBRL algorithms learn a dynamics model from these interactions first. Despite the success of MFRL in several domains as robotics and video games \cite{Heess2017, Schulman2017,Silver2014,Lillicrap2015,OpenAI2018}, their applications is mainly restricted to simulated environments due to their low sample-efficiency. However, putting constraints on the algorithms, as discussed in \cref{s:Model-free reinforcement learning}, partly cures the sample inefficiency, turning them into practical tools for real-world problems.\\
	On the other hand, MBRL showing a much better sample efficiency. The challenge for these methods is to learn an accurate model, which turns out to be rather hard and leads to the so-called model-bias problem \cite{Deisenroth2011}. In \cref{ss:Uncertainty aware DYNA-style reinforcement learning} we discuss attempts to capture the model uncertainty to alleviate this obstacle by making use of model ensembles and Bayesian neural networks resulting in algorithms, exhibiting a high sample-efficiency and a low model-bias.\\
	In what follows, we denote the set of all states $\bs$ by $\states$ and the set of all actions $\ba$ by $\actions$. Relative to state and action vectors, $(\bs,\ba)$, we define a reward function $\reward : \states \times \actions \rightarrow \real$, a scalar $\discount \in (0,1]$ called discount factor and an initial state distribution $\initstate$. $\transitions(\bs_{t+1}|\bs_t,\ba_t)$ characterises the probability to end in a state $\bs_{t+1}$ if an action $\ba_t$ is taken at state $\bs_t$. The tuple \mbox{$\mdp = (\states,\actions,\transitions,\initstate,\reward,\discount)$} defines a Markov decision process.\\
	Goal of reinforcement learning is to find a $\policy^*: \bs\mapsto \ba$, which is the solution of:
	\begin{align}
		J(\policy^*)=\max J (\policy)  = 
		\mathbb E_{\traj \sim p_{\policy}}\left[\sum_{t=0}^{\horizon}\discount^t r(\bs_t,\ba_t)\right],\label{eq:cumulative_reward}
	\end{align}
	where
	\begin{align}
		p_\policy = \initstate(\bs_0)\prod_{t_0}^\horizon\policy(\ba_t,\bs_t)\transitions(\bs_{t+1}|\bs_t,\ba_t)
		\label{eq:trajectory_distribution}
	\end{align}

@@

	In the first case $\policy$ is optimized directly and one speaks about policy gradient methods \cite{Sutton2018, Williams1992,Baxter2011,pmlr-v28-levine13, Schulman2015,Schulman2017}. To train these methods, a large number of direct interactions with the system is required, since they belong to \emph{on-policy} algorithms. Hence, we rule them out for direct online training but make use of their characteristics in MBRL (\cref{ss:critical_design}).\\

@@

	As already adduced, MFRL algorithms learn directly from interactions with the system. An exhaustive overview can be found, e.g. in \cite{Sutton2018,Levine2020}. In the following, we discuss the normalised advantage function (\emph{NAF}) algorithm \cite{Gu2016} in some detail, as it has good characteristics for accelerator control problems. It is highly sample-efficiency, and its representational power is sufficient for most common cases\cite{Kain2020,Hirlaender2020a}. Modifications to increase the stability are subject of \cref{ss:Normalized advantage function}.

@@

	\Cref{tab:overview_algorithms} shows a comparison of the discussed algorithms. Details are provided in the appendix. All algorithms are suitable for control tasks as common in accelerator operation. Their strength lies in the high sample-efficiency, but they differ in representational power and noise robustness. In \cref{sec:Experimental results from FERMI RL online tests} the \emph{NAF2},

@@

	\\ To exploit the convergence properties of the \emph{TRPO}, the policy was never reset as can be seen in \cref{fig:ME-TRPO_observables}. In the upper figure, the total reward per batch gathered from the real environment, and the number of data-points used to train the dynamics model as a function of the number of epochs is shown. In the lower plot, the average cumulative reward of ten episodes as achieved by the \emph{TRPO} on the individual models of the ensemble independence of the epochs is drawn. During an epoch the \emph{TRPO} is trained for 10000 steps on the synthetic data. The shaded area shows the corresponding standard deviation to indicate the uncertainty of the dynamics model. After initial performance drops caused by the insufficiently trained model, the algorithm continuously improves. As a measure of convergence of the \emph{TRPO}, the logarithm of the standard deviation of $p_\policy$ (\cref{eq:trajectory_distribution}) is visualised. The training was stopped after 450 steps collecting 25 steps each dynamics training.

@@

	In contrast to the \emph{ME-TRPO} training, the data batches consisted of 50 data-points with an initial random walk of 200 steps. The number of initial steps was chosen high enough because otherwise the convergence is slowed down strongly so that the training becomes unfeasible on a real machine (discussed in \cref{ss:critical_design}). Each epoch consists of 2500 steps of controller training on the model. The training was stopped after the acquisition of 500 data-points. The verification was executed as in the first test. Again, the success of all 50 episodes is 100\%. The number of needed iterations per episodes is less than for the ME-TRPO, which can be seen in \cref{fig:AE-DYNA_verification}. The somewhat better performance might be a result of the higher number of data-points (50), but in general, this method exhibited better asymptotic performance than the \emph{ME-TRPO} variant.\\

@@

	Both experiments yielded satisfactory results showing that a non-linear and noisy problem could be solved in a feasible number of training steps. The results of the verification are summarised in \cref{tab:overview_verification}. In the model-free case, around 800-1000 and in the model-based 450-500 data-points were used. In all experiments, all verification episodes were finished successfully. \emph{NAF2} performs better in terms of episode length while the average reward in the \emph{DYNA}-style algorithms is higher. The experiments were done on different days, hence under slightly different conditions.\\

@@

	The MFRL methods were slightly better in the final performance, but also more samples have been collected during the training. The MBRL methods had to be stopped due to lack of available experimental time. Hence additional studies regarding a long time performance would be interesting.\\ One big issue in applying MBRL methods can be the computational time to train the controller on the model. In our tests, the data acquisition time was only a fraction of the time needed for the controller-training. In case, the used methods could be parallelised to reduce the computational time.

@@

	The presented reinforcement learning methods hold tremendous promise for automatizing set-ups typical in accelerators. One objective of this work was to provide some suggestions on how to make advanced deep reinforcement learning techniques work on a real set-up by adapting available methods. This was demonstrated on the FERMI FEL intensity optimization problem.\\

@@

	To provide the possibility for other laboratories to profit form the stated methods, the code was released in \url{https://github.com/MathPhysSim/FERMI_RL_Paper} and \cite{Hirlaender2020b}.

>>>>

	\begin{abstract}
		Reinforcement learning holds significant potential in accelerator controls. The primary goal of this paper is to demonstrate the operational application of this approach to accelerator physics problems. Despite the success of model-free reinforcement learning in several domains, sample efficiency remains a limiting factor, a challenge that model-based methods may mitigate.
		We compare well-suited purely model-based to model-free reinforcement learning applied to the intensity optimisation on the FERMI FEL system. We find that the model-based approach demonstrates higher representational power and sample-efficiency, while the asymptotic performance of the model-free method is slightly superior. The model-based algorithm is implemented in a DYNA-style using an uncertainty aware model, and the model-free algorithm is based on tailored deep Q-learning. In both cases, the algorithms were implemented in a way, which presents increased noise robustness as prevalent in accelerator control problems. Code is released in \url{https://github.com/MathPhysSim/FERMI_RL_Paper}.

@@

	In particle accelerator operation, one main goal is to provide stable and reproducible performance. Achieving this requires addressing several control problems simultaneously \cite{Kain2020,Scheinker2018}. Particularly when a priori physics modeling is infeasible, one may employ optimisation techniques such as derivative-free optimisers (DFOs) \cite{Huang2013,Bruchon2017,Scheinker2020,Hirlaender2019,Welsch2015,Albright2019} or model-based optimisations as Gaussian Processes \cite{Hanuka2020,Roussel2020} to restore or maintain the performance.\\

@@

		\item It retains information about the problem structure and does not always begin from zero as a DFO.

@@

	One demanding aspect using RL in real-world applications is the number of iterations needed to train a controller - the sample efficiency, since RL methods are known to be data-intensive \cite{Sutton2018,DulacArnold2019}. A second critical aspect to be considered is the robustness of training in a noisy, data-limited environment.\\

@@

	We successfully applied two different algorithm classes, a highly sample-efficient variant of continuous Q-learning (\cref{ss:Normalized advantage function}) and a newly proposed model-based algorithm, which we call \emph{AE-DYNA} (\cref{ss:Uncertainty aware DYNA-style reinforcement learning}), to the FERMI FEL problem.\\ The methods exhibit distinct characteristics in addressing the aforementioned critical aspects. They are generally applicable to problems showing a sample-complexity as in common accelerator control problems.

@@

	 Thus, we demonstrate the operational applicability of purely model-based reinforcement learning to an accelerator control problem.\\

@@

		\item Description of the problem setup at FERMI

@@

	\section{The setup of the studied problem}
	\subsection{The Physical Setup}

@@

		\caption{A schematic view on the setup of the FERMI FEL.}

@@

	A schematic overview of the setup is provided in \cref{fig:schematic_FEL}.

@@

	The state $\bs$ is a four dimensional vector holding the current voltage values applied to the piezo motors, where each component lies within the normalized interval [0, 1]. Two values correspond to the two DOF of each mirror. The action $\ba$ is four-dimensional, namely a delta step on the voltages at time-step $t$ $\bs_t$:

@@

	\section{Deep Reinforcement learning}
	Most commonly, two main categories of deep RL algorithms are distinguished: model-free RL (MFRL) and model-based RL (MBRL). MFRL algorithms learn directly from the interaction with a system, while MBRL algorithms learn a dynamics model from these interactions first. Despite the success of MFRL in several domains as robotics and video games \cite{Heess2017, Schulman2017,Silver2014,Lillicrap2015,OpenAI2018}, their applications is mainly restricted to simulated environments due to their low sample-efficiency. However, putting constraints on the algorithms, as discussed in \cref{s:Model-free reinforcement learning}, partly cures the sample inefficiency, turning them into practical tools for real-world problems.\\
	On the other hand, MBRL shows much better sample efficiency. The challenge for these methods is to learn an accurate model, which turns out to be rather hard and leads to the so-called model-bias problem \cite{Deisenroth2011}. In \cref{ss:Uncertainty aware DYNA-style reinforcement learning} we discuss attempts to capture the model uncertainty to alleviate this obstacle by making use of model ensembles and Bayesian neural networks resulting in algorithms, exhibiting a high sample efficiency and a low model-bias.\\
	In what follows, we denote the set of all states $\bs$ by $\states$ and the set of all actions $\ba$ by $\actions$. Relative to state and action vectors, $(\bs,\ba)$, we define a reward function $\reward : \states \times \actions \rightarrow \real$, a scalar $\discount \in (0,1]$ called discount factor and an initial state distribution $\initstate$. $\transitions(\bs_{t+1}|\bs_t,\ba_t)$ characterises the probability to end in a state $\bs_{t+1}$ if an action $\ba_t$ is taken at state $\bs_t$. The tuple \mbox{$\mdp = (\states,\actions,\transitions,\initstate,\reward,\discount)$} defines a Markov decision process.\\
	The goal of reinforcement learning is to find a $\policy^*: \bs\mapsto \ba$, which is the solution of:
	\begin{align}
		J(\policy^*)=\max J (\policy)  = 
		\mathbb E_{\traj \sim p_{\policy}}\left[\sum_{t=0}^{\horizon}\discount^t r(\bs_t,\ba_t)\right],\label{eq:cumulative_reward}
	\end{align}
	where
	\begin{align}
		p_\policy = \initstate(\bs_0)\prod_{t_0}^\horizon\policy(\ba_t,\bs_t)\transitions(\bs_{t+1}|\bs_t,\ba_t)
		\label{eq:trajectory_distribution}
	\end{align}

@@

	In the first case $\policy$ is optimised directly and one speaks about policy gradient methods \cite{Sutton2018, Williams1992,Baxter2011,pmlr-v28-levine13, Schulman2015,Schulman2017}. To train these methods, a large number of direct interactions with the system is required, since they belong to \emph{on-policy} algorithms. Hence, we rule them out for direct online training but make use of their characteristics in MBRL (\cref{ss:critical_design}).\\

@@

	As already adduced, MFRL algorithms learn directly from interactions with the system. An exhaustive overview can be found, e.g. in \cite{Sutton2018,Levine2020}. In the following, we discuss the normalised advantage function (\emph{NAF}) algorithm \cite{Gu2016} in some detail, as it has good characteristics for accelerator control problems. It has high sample efficiency, and its representational power is sufficient for most common cases\cite{Kain2020,Hirlaender2020a}. Modifications to increase the stability are subject of \cref{ss:Normalized advantage function}.

@@

	\Cref{tab:overview_algorithms} shows a comparison of the discussed algorithms. Details are provided in the appendix. All algorithms are suitable for control tasks as common in accelerator operation. Their strength lies in the high sample efficiency, but they differ in representational power and noise robustness. In \cref{sec:Experimental results from FERMI RL online tests}, the \emph{NAF2},

@@

	\\ To exploit the convergence properties of the \emph{TRPO}, the policy was never reset as can be seen in \cref{fig:ME-TRPO_observables}. In the upper figure, the total reward per batch gathered from the real environment, and the number of data-points used to train the dynamics model as a function of the number of epochs is shown. In the lower plot, the average cumulative reward of ten episodes as achieved by the \emph{TRPO} on the individual models of the ensemble independently of the epochs is drawn. During an epoch the \emph{TRPO} is trained for 10000 steps on the synthetic data. The shaded area shows the corresponding standard deviation to indicate the uncertainty of the dynamics model. After initial performance drops caused by the insufficiently trained model, the algorithm continuously improves. As a measure of convergence of the \emph{TRPO}, the logarithm of the standard deviation of $p_\policy$ (\cref{eq:trajectory_distribution}) is visualised. The training was stopped after 450 steps collecting 25 steps each dynamics training.

@@

	In contrast to the \emph{ME-TRPO} training, the data batches consisted of 50 data-points with an initial random walk of 200 steps. The number of initial steps was chosen high enough because otherwise the convergence is slowed down strongly so that the training becomes unfeasible on a real machine (discussed in \cref{ss:critical_design}). Each epoch consists of 2500 steps of controller training on the model. The training was stopped after the acquisition of 500 data-points. The verification was executed as in the first test. Again, the success rate of all 50 episodes is 100\%. The number of needed iterations per episodes is less than for the ME-TRPO, which can be seen in \cref{fig:AE-DYNA_verification}. The somewhat better performance might be a result of the higher number of data-points (50), but in general, this method exhibited better asymptotic performance than the \emph{ME-TRPO} variant.\\

@@

	Both experiments yielded satisfactory results, demonstrating that a non-linear, noisy problem can be solved within a feasible number of training steps. The results of the verification are summarised in \cref{tab:overview_verification}. In the model-free case, around 800-1000 and in the model-based 450-500 data-points were used. In all experiments, all verification episodes were finished successfully. \emph{NAF2} performs better in terms of episode length while the average reward in the \emph{DYNA}-style algorithms is higher. The experiments were done on different days, hence under slightly different conditions.\\

@@

	The MFRL methods were slightly better in the final performance, but also more samples have been collected during the training. The MBRL methods had to be stopped due to lack of available experimental time. Hence additional studies regarding a long time performance would be interesting.\\ A significant challenge in applying MBRL methods is the computational time to train the controller on the model. In our tests, the data acquisition time was only a fraction of the time needed for the controller-training. In case, the used methods could be parallelised to reduce the computational time.

@@

	The presented reinforcement learning methods hold tremendous promise for automating setups typical in accelerators. One objective of this work was to provide some suggestions on how to make advanced deep reinforcement learning techniques work on a real setup by adapting available methods. This was demonstrated on the FERMI FEL intensity optimisation problem.\\

@@

	To provide the possibility for other laboratories to profit from the stated methods, the code was released in \url{https://github.com/MathPhysSim/FERMI_RL_Paper} and \cite{Hirlaender2020b}.
